---
title: "GapMinder_Analysis"
author: "Prateek Masali"
date: "2/10/2019"
output:
  html_document:
    toc: true
    theme: cosmo
    highlight: tango
    df_print: paged
    code_folding: hide
    hr { padding: 2px 0 2px 0; }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## Question
*To measure the effects of Geographic location, Income or Population on life expectance rate*

---

### Data Description
- Region/Country fields describes the continent and country the Life expectancy is measured
- Additionally, Population, Income, and Life expectance is provided for each country from years 1800 to 2015

###### *Income - Gross domestic product per person adjusted for differences in purchasing power (GDP/capita, PPP$ inflation adjusted)*
###### *Life - The average number of years a newborn child would live if  current mortality pattern were to stay the same*
###### *population - Population for the given country in the given year  [(Source: GapMinder)](https://www.gapminder.org/data/)*

### Data Preparation
1. Import the Raw data
2. Summarize Data
3. Study the scope

```{r gapminder}
#Importing raw data
gapminder <- read.csv("~/Desktop/Harrisburg University/ANLY 506-90-O/Exploratory-Data-Analytics/Data/gapminder.csv")
#Summary of data
str(gapminder)
```
Here, we study that Country, population, and income are Factors. Life and income are of num data type while year is of int datatype.
```{r}
#Study the scope
summary(gapminder)
levels(gapminder$Country)
```
Looking at the summary, now we know there are 197 countries spread across 6 regions, each having 216 entries for years from 1800 to 2015. The life expectancy ranges from 1 to 84%

---

## Analysis

```{r}

#gapminder_2015 <- gapminder %>% filter(Year==2015)
#symbols(gapminder_2015$income, gapminder_2015$life, circles = gapminder_2015$population, inches=0.35, fg="white", bg="red", xlab="Income", ylab="Life Expectancy") + scale_x_log10()
#text(gapminder_2015$income, gapminder_2015$life, gapminder_2015$Country, cex=0.4)


```

```{r}

library("plotly")
p <- gapminder %>%
              filter(Year==2015) %>%
                  ggplot(aes(income, life, size = population, color=region)) + geom_point() + scale_x_log10() +
                  theme_bw()
 
ggplotly(p)

```

## Including Plots

You can also embed plots, for example:

```{r income, echo=FALSE}
plot(hist(gapminder$life))
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
